Diagnosising the sensitivity of grounding line flux to changes in sub-ice shelf melting

### 2019-04-23 draft_v1.tex
put new fig and modify/add some paragraphs

@@

%\subsection{Possible controlling factors}
\subsection{Understanding the dependence on buttressing direction}

@@

\begin{figure}
	\centering
    \includegraphics[width=1\linewidth]{figs/fig4.pdf}
    \caption{Histograms for the frequency of the maximum (red) and minimum (blue) velocity changes in neighboring grid cells as a function of angular distance around each perturbation point. In a and c, all perturbation points are shown and in plots b and d, only the subset of points from Figure \ref{fig2}d (b and d) are shown. ({\bf{TZ: this figure still needs double check!}})}
	\label{fig4}
	\label{fig4}
\end{figure}

Further evidence for the importance of the first principle stress direction is given by the variances that characterize the changes in the state of stress and the state of velocity between the initial condition and the perturbation experiments. To this end, we define the metric $\Psi$,  
\begin{equation}
\Psi = \frac{\sigma_{p}-\sigma_{c}}{\sigma_{c}} - \frac{u_{p}-u_{c}}{u_{c}},
\label{psi}
\end{equation}

where the subscripts $p$ and $c$ denote the perturbation experiments and the ``control'' (i.e., the initial condition), respectively, and $\sigma$ and $u$ denote the stress component and ice velocity, respectively. This metric is a measure of the consistency between changes in the model stress and velocity states between the control (the initial condition) and perturbation experiments. We calculate $\Psi$ at every grid point for the case where  $\sigma_{nn}$ in Equation \ref{butN} is set equal to $\sigma_{p1}$ (Fig \ref{fig5}a) and $\sigma_{p2}$ (Fig \ref{fig5}b). From Figure \ref{fig5}, it is clear that the variance between stress and velocity change is much smaller everywhere in the domain, including along the GL, for the case of $\sigma_{nn}=\sigma_{p1}$. Physically this can be interpreted as demonstrating a stronger correlation between changes in the first principle stress component and the velocity (or flux) than between changes in the second principle stress. \textit{SP: Or something like this. Need to think about it some more still. I wonder if it could also be interpreted differently -- e.g., in 5b, could the much larger variance be because a small change in p2 results in a much larger change in velocity (relative to p1)?. Would it make more sense for Fig. 5 to be an actual correlation coeff. plot rather than the somewhat complicated variance comparison?}. Examination of the stress and velocity changes at the GL for a perturbation at a specific location on the ice shelf provides further evidence of a strong (weak) correlation between changes in the first principle stress (second principle stress) and changes in the velocity (Figure \ref{fig6}).

@@

\begin{figure}
	\centering
    \includegraphics[width=1\linewidth]{figs/fig5.pdf}
    \caption{The standard deviations of $\Psi$ -- a metric for how closely a change in normal stress corresponds with a change in velocity -- at each perturbation point of the MISMIP+ domain, for the case of $\sigma_{nn}=\sigma_{p1}$ (a) and $\sigma_{nn}=\sigma_{p2}$ (b).}
	\label{fig5}
\end{figure}

\textit{SP: Leaving this section alone for now. I think it's worth including some additional discussion here if we can make some reasonable speculation as to what's going on. But we probably need to come up with a bit more than this. I wonder if there are other T3 folks who might have some insight into this problem? One thing that occurs to me is that we could be conflating cause and effect here. For example, when we look at Figure 6, it looks like the changes in velocity correlate with changes in $\sigma_{p1}$, whereas they don't correlate with changes in $\sigma_{p2}$. But it could simply be that the changes in $\sigma_{p2}$ (due to the thickness perturbation) are causing a velocity change, and the velocity change is better reflected by the change in $\sigma_{p1}$. This makes sense at the GL, where we know the velocity / flux are largely going to be aligned with the $\sigma_{p1}$ direction (perpendicular to the GL). This doesn't, however, explain why we see a good correlation between changes in velocity / flux at the GL (reflected by the flux response number, N) and $\sigma_{p1}$ but a much poorer correlation between the flux response number and $\sigma_{p2}$.} One of the possible reasons for us seeing such direction-dependent correlations might be due to the perturbation propagation features on ice shelves. The energy of perturbation propagates with the group velocity if we decompose it using Fourier transform. A similar example can be found in \citep{gudmundsson2003}. Using very simplified geometry \cite{gudmundsson2003} analyzed the propagation of basal perturbation to glacier surface and found that the direction of group velocity is aligned closely with the main flow. The existence of preferred propagation direction for the perturbations can possibly lead to our findings that favor the first principle stresses ({\bf{TZ: I am still not able to explain why it's exactly the first principle stress direction :(}}).

@@

@@

\begin{equation}
    \Lambda = \sum_i^I \frac{1}{d_i}\mid cos\theta_i \mid,
    \ref{Lambda}
\end{equation}
where $I$ is the total cell number on GL, $d_i$ is the distance between the perturbation spot and the $i$-th cell on GL, $theta_i$ is the angle difference between the direction of the specified normal stress at the perturbation spot and the flow direction for the $i$-th cell on GL. This metric gives us two possible important factors that impact GLF: 1) the geometric location of perturbation spot(the closer perturbation spots to GL, the larger impact of perturbation on GLF; see the discussion in the following section ``Impacts of near-GL perturbations''), 2) the relationship between ice flow along GL and the buttressing direction we choose at the perturbation spot (the perturbation has more impacts on GLF if the buttressing direction aligned more closely with the ice flow directions on GL).  
Similar to above, we rotate the buttressing direction by 1--180$^\circ$ on top of the $\sigma_{p1}$ direction, and calculate $\Lambda$ for each perturbation spot. From Figure \ref{new_metric_new} we can see that $\Lambda$ has large values when the buttressing directions are close to the $\sigma_{p1}$ direction ($\Delta\phi=0^\circ$ or $\Delta\phi=180^\circ$). This is because, for MISMIP+, the $\sigma+{p1}$ on GL largely controls the GLF (see the quiver plots of principle stresses in the Appendix). For real ice shelves, the pattern of ice flow along GL might be much more complex then of MISMIP+m, resulting in difficulty of predicting GLF using single direction buttressing number.

>>>>

In the following sections, we elaborate our understandings of the dependence on buttressing directions.

@@

%\subsection{Possible controlling factors}
%\subsection{Understanding the dependence on buttressing direction}
\subsection{The changes of velocity around the perturbation spot}

@@

\begin{figure}
	\centering
    \includegraphics[width=1\linewidth]{figs/fig4.pdf}
    \caption{Histograms for the frequency of the maximum (red) and minimum (blue) velocity changes in neighboring grid cells as a function of angular distance around each perturbation point. In a and c, all perturbation points are shown and in plots b and d, only the subset of points from Figure \ref{fig2}d (b and d) are shown. ({\bf{TZ: this figure still needs double check!}})}
	\label{fig4}
\end{figure}

\subsection{The changes of velocity and stress along GL}

Further evidence for the importance of the first principle stress direction is given by the variances that characterize the changes in the state of stress and the state of velocity between the initial condition and the perturbation experiments. To this end, we define the metric $\Upsilon$,  
\begin{equation}
    \Upsilon= corr\left(\frac{\sigma_{p}-\sigma_{c}}{\sigma_{c}}, \frac{u_{p}-u_{c}}{u_{c}}\right),
\label{upsilon}
\end{equation}

%\begin{equation}
%    \Psi= \frac{\sigma_{p}-\sigma_{c}}{\sigma_{c}}-\frac{u_{p}-u_{c}}{u_{c}},
%\label{psi}
%\end{equation}
where the subscripts $p$ and $c$ denote the perturbation experiments and the ``control'' (i.e., the initial condition), respectively, and $\sigma$ and $u$ denote the stress component and ice velocity, respectively. Here the changes of $\Upsilon$ we discuss is limited to GL. $\Upsilon$ are a measure of the consistency between changes in the model stress and velocity states between the control (the initial condition) and perturbation experiments. We calculate $\Upsilon$ for every perturbation point for each direction from 0--180$^\circ$ (Figure \ref{stress_vel_corr_GL_allP}). Clearly, when we pick the $\sigma_{p1}$ direction as the buttressing direction, we can find the best correlation between the changes of ice surface speed and normal stress along GL, whereas the correlation numbers are relatively low when the buttressing directions are close to the $\sigma_{p2}$ direction.  
\begin{figure}
	\centering
    \includegraphics[width=1\linewidth]{figs/stress_vel_corr_GL_allP.pdf}
    \caption{The correlation ($\Upsilon$) between the changes of ice surface speed and the changes of normal stress along GL. The direction of normal stress is rotated anti-clockwisely from the direction of $\sigma_{p1}$. Each colored curve represents a perturbation experiment, and the thick dashed back curve is their mean value.}
	\label{stress_vel_corr_GL_allP}
\end{figure}

More specifically, we can see the spatial pattern of $\Upsilon$ for every perturbation point for the case where $\sigma_{nn}$ in Equation \ref{butN} is set equal to $\sigma_{p1}$ (Fig \ref{fig5}a) and $\sigma_{p2}$ (Fig \ref{fig5}b). From Figure \ref{fig5}, it is clear that the correlation between stress and velocity change along GL is generally much larger in the domain for the case of $\sigma_{nn}=\sigma_{p1}$, especially for some regions in the center of the domain. \textit{SP: Or something like this. Need to think about it some more still. I wonder if it could also be interpreted differently -- e.g., in 5b, could the much larger variance be because a small change in p2 results in a much larger change in velocity (relative to p1)?. Would it make more sense for Fig. 5 to be an actual correlation coeff. plot rather than the somewhat complicated variance comparison?}. \textbf{TZ: I now change deviation to correlation. You are right. It's more consistent. I think the problem is to check if the changes of $\sigma_{p1}$ or $\sigma_{p2}$ can cause more consistent velocity changes along GL, right? For example, if $\sigma_{p2}$ can cause a larger but more consistent change in vel. than $\sigma_{p1}$, then $\sigma_{p2}$ is still the better metric than $\sigma_{p1}$.} In Figure \ref{fig6} we show the stress and velocity changes at the GL for a perturbation at a specific location on the ice shelf, which may provide a more clear evidence of a strong (weak) correlation between changes in the first principle stress (second principle stress) and changes in the velocity (Figure \ref{fig6}). For this particular perturbation spot, we can also see the changes of buttressing number calculated by $\sigma_{p1}$ (Figure \ref{sigma_change_example}a) and $\sigma_{p2}$ (Figure \ref{sigma_change_example}b) across the domain. In this case the buttressing number calculated from $\sigma_{p1}$ shows more decrease than that from $\sigma_{p2}$ for most upstream parts to the perturbation spot, indicating another evidence that the more tensile stress $\sigma_{p1}$ can change more significantly and therefore contribute more to the changes of ice shelf buttressing under certain basal melt forcing scenarios.

@@

\begin{figure}
	\centering
    \includegraphics[width=1\linewidth]{figs/stressDiff_velDiff_GL_allPerturb.pdf}
    \caption{The standard deviations of $\Psi$ -- a metric for how closely a change in normal stress corresponds with a change in velocity -- at each perturbation point of the MISMIP+ domain, for the case of $\sigma_{nn}=\sigma_{p1}$ (a) and $\sigma_{nn}=\sigma_{p2}$ (b).}
	\label{fig5}
\end{figure}

\textit{SP: Leaving this section alone for now. I think it's worth including some additional discussion here if we can make some reasonable speculation as to what's going on \textbf{TZ: The metric regarding distance and flow/stress angle is a such speculation that the impact of perturbation on GLF is a combined effect of perturbation and GL dynamics (see below)}. But we probably need to come up with a bit more than this. I wonder if there are other T3 folks who might have some insight into this problem \textbf{TZ: there might be some similar theory going on here, but I doubt the pure mechanical knowledge apply in this specific problem. Perhaps Xylar can contribute a bit from the perspective of Fourier transform theory} ? One thing that occurs to me is that we could be conflating cause and effect here. For example, when we look at Figure 6, it looks like the changes in velocity correlate with changes in $\sigma_{p1}$, whereas they don't correlate with changes in $\sigma_{p2}$. But it could simply be that the changes in $\sigma_{p2}$ (due to the thickness perturbation) are causing a velocity change, and the velocity change is better reflected by the change in $\sigma_{p1}$. This makes sense at the GL, where we know the velocity / flux are largely going to be aligned with the $\sigma_{p1}$ direction (perpendicular to the GL). This doesn't, however, explain why we see a good correlation between changes in velocity / flux at the GL (reflected by the flux response number, N) and $\sigma_{p1}$ but a much poorer correlation between the flux response number and $\sigma_{p2}$. \textbf{TZ: I am not sure if I follow you here. I don't know if it's a proper way to say that the changes in $\sigma_{p2}$ cause the changes in velocity. If we look closely at Figure \ref{fig6} we can see that for some cells on GL, even though the velocity direction is more aligned with p1, the correlation is still poor, whereas the correlation with p2 is better for these same cells. The difference here is their respective angle difference with the stress direction at perturbation spot (a main reason I make up the metric in the following; I can show you some point when you have time). It's hard to know exactly the reason behind.}} One of the possible reasons for us seeing such direction-dependent correlations might be due to the perturbation propagation features on ice shelves. The energy of perturbation propagates with the group velocity if we decompose it using Fourier transform. A similar example can be found in \citep{gudmundsson2003}. Using very simplified geometry \cite{gudmundsson2003} analyzed the propagation of basal perturbation to glacier surface and found that the direction of group velocity is aligned closely with the main flow. The existence of preferred propagation direction for the perturbations can possibly lead to our findings that favor the first principle stresses (\textbf{TZ: I am still not able to explain why it's exactly the first principle stress direction :(}).

@@

\begin{figure}
	\centering
    \includegraphics[width=1\linewidth]{figs/sigma_change_example.pdf}
    \caption{Relationship between GL {changes} in the velocity (blue) and in the stress (red) for the MISMIP+ test case due to a perturbation at a specific location on the ice shelf (blue dot in inset map). In a), changes in GL velocity are plotted against changes in $\sigma_{p1}$ and in b) changes in GL velocity are plotted against changes in $\sigma_{p2}$. The $x$-axis is an index for the grid cell number along the GL.}
	\label{sigma_change_example}
\end{figure}

@@

\begin{equation}
    \Lambda = \sum_i^I \frac{1}{d_i} |\mathrm{cos} \left(\theta_i\right)|,
    \label{Lambda}
\end{equation}
where $I$ is the total cell number on GL, $d_i$ is the distance between the perturbation spot and the $i$-th cell on GL, $theta_i$ is the angle difference between the direction of the specified normal stress at the perturbation spot and the flow direction for the $i$-th cell on GL. This metric gives us two possible important factors that impact GLF: 1) the geometric location of perturbation spot(the closer perturbation spots to GL, the larger impact of perturbation on GLF; see the discussion in the following section ``Impacts of near-GL perturbations''), 2) the relationship between ice flow along GL and the buttressing direction we choose at the perturbation spot (the perturbation has more impacts on GLF if the buttressing direction aligned more closely with the ice flow directions on GL).  
Similar to above, we rotate the buttressing direction by 1--180$^\circ$ on top of the $\sigma_{p1}$ direction, and calculate $\Lambda$ for each perturbation spot. From Figure \ref{new_metric_new} we can see that $\Lambda$ has large values when the buttressing directions are close to the $\sigma_{p1}$ direction ($\Delta\phi=0^\circ$ or $\Delta\phi=180^\circ$). This is because, for MISMIP+, the $\sigma+{p1}$ on GL largely controls the GLF (see the quiver plots of principle stresses in the Appendix). For real ice shelves, the pattern of ice flow along GL might be much more complex then of MISMIP+m, resulting in difficulty of predicting GLF using single direction buttressing number.
